Disentangling patterns of gene expression in high grade serous ovarian cancer by matrix factorization

### 2020-08-08 MSc_Dissertation_B156476.tex
Substantial work on 'Epithelial Ovarian Cancer' and 'Patterns of ... in HGSOC'

\documentclass[tikz, 12pt,a4paper,oneside,fleqn]{article}
\usepackage[margin=1.5cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{setspace}
\usepackage{enumitem}
\setlist[description]{leftmargin=\parindent,labelindent=\parindent}

@@

\title{Disentangling patterns of gene expression in high grade serous ovarian cancer}
\author{}

@@

%%%%%% Assesment Title Page
\begin{center}
\includegraphics[scale=0.3]{images/UoE_SBO_logo.png}
\end{center}

@@

Disssertation Supervisor:  Dr. Ailith Ewing
\vspace{0.2in}
\end{center}
\end{mdframed}

@@

\section{Introduction}
High grade serous ovarian cancer (HGSOC) effects the epithelium of the ovaries and fallopian tubes.   ``Serous" refers to the epithelial membrane which secretes serous fluid (serum).  It is the 7th most common cancer in women worldwide .  Epithelial ovarian cancers are classified into two broad subtypes:  Type I are low grade with generally good prognosis; type II are more aggressive and typically carry P53 mutations with defects in mechanisms of DNA repair.  Type II is more-or-less synonomous with HGSOC. \cite{Lisio2019}.

This work is concerned with analysing the patters gene expression in HGSOC by dimensionality reduction techniques of matrix factorization, and relating these patterns to clinical and genomic features.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.4, draft=false]{Sompairac_ICA_NMF_PCA.png}
\end{center}
\caption{Illustrating the differing results of component extraction by PCA, ICA and NMF.  From \cite{Sompairac2019}}
\label{fig-pca-ica-nmp-explanation}
\end{figure}

@@

In this work we wish to treat and discuss the three factorization methods in a unified way.  Thus, for the remainder of this dissertation we adopt the following conventions:\footnote{I should probably use bold font for X, W and H matrices}.
\begin{description}
\item[Expression matrix:] $X$ $(M \times N)$
\item[Metagene matrix:] $W$ $(M \times K)$
\item[Metasample matrix:] $H$ $(K \times N)$
\end{description}
where $N$ is the number of patients (or samples), $M$ is the number of genes and $K$ is the factorization rank, i.e. the number of components or factors.  Thus the factorization is written
\begin{equation}
X \approx W H
\end{equation}
Adding subscripts to indicate the row,column orientation of the matrices:
\begin{equation}
X_{M,N} \approx W_{M,K} H_{K,N}
\end{equation}
as illustrated in figure \ref{fig-matrix-factorization} (a).

@@

\subsection{Gene expression in HGSOC}

This section focusses on biological and clinical conclusions relating to gene expression in HGSOC and mentions analysis techniques only in passing.

Obesity is known to negatively impact prognosis in ovarian cancer \cite{Cuello2018}, \cite{Au-Yeung2014}.  Cuello \etal\ \cite{Cuello2018} used NMF based clustering on mRNA microarrays and reverse-phase protein arrays to demonstrate an association between cancer driver genes and obesity related genes.

Note that several of the studies reviewed here use data from the Australian Ovarian Cancer Study (AOCS): \cite{Winterhoff2016,Patch2015,Ewing2020,Zhang2018,Cuello2018,Au-Yeung2014}.

Stratification of HGSOC by immunohistochemistry is addresses Kamble \etal. \cite{Kamble2019}.  They identify a panel of six biomarkers -- TCF21, E-cadherin, PARP1, Slug and AnnexinA2.   These markers allow stratification into categories including Mesenchymal-to-epithelial transition, Homologous recombination repair and Epithelial-to-mesenchymal transition.  This, they argue, allows selection of class-specific inhibitor drugs such as Olaparib and Rucaparib.

\subsection{Research questions}
The following research goals were set out early in project. 
\begin{enumerate}
\item Which are the influential genes in HGSOC according to gene expression data?   Do these confirm published results?
\item Which of PCA, ICA and NMF is best suited to this analysis?
\item For identified genes, what is the underlying biology?
\item Can we discriminate subtypes of HGSOC from expression data?
\item Is there value in incorporating somatic and germline genomic data?
\end{enumerate}

@@

Our overall approach is to use \emph{unsupervised} machine learning methods to represent gene expression data in a small number of features, then to study whether these features correlate with clinical and biological variables.  This is appropriate in our case, since of the two datasets available (described below), one (TCGA) has relatively large $n$ but little available metadata, while the other (AOCS) is much smaller but has useful metadata.   This motivates an approach of unsupervised learning on the TCGA dataset.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.7, draft=false]{methodlology_overview.pdf}
\end{center}
\caption{Overview of methodology as information flow.  The diagramatic convention here is that at each stage, both datasets -- TCGA and AOCS -- are processed separately by the given algorithm.  The exception is Gene Set Intersection, which involves interaction between the datasets.}
\label{fig-method-overview}
\end{figure}
In outline our method is follows:
\begin{enumerate}
\item
Identification of a consistent gene set between the TCGA and AOCS datasets.
\item
Unsupervised metagene extraction by matrix factorization methods : NMF, ICA and PCA.
\item
Metagene selection for robustness by k-means clustering of bootstrap resampled factorizations evaluated by silhouette score, based on the TCGA dataset.
\item
Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO).
\item
Computation by least squares optimisation of metasamples associated with with each TCGA derived metagene against TCGA and AOCS datasets.
\item
Alignment of available metadata with computed metasamples, for TCGA and AOCS.
\item 
Survival analysis on TCGA and AOCS to investigate relationship between metasamples and patient survival.
\item
Clustered heatmaps to visualise relationships between metasamples, patients and metadata.
\end{enumerate}

@@

       Ws_reduced = PCA(Ws, rank=20)
       plot t-SNE(Ws_reduced)
       clustering = k-means clustering(Ws_reduced, n_clusters=k)
       score[factorizer, k] = silhouette score(clustering)	  
       median_metagenes = calculate medians for k clusters of metagenes
       save median_metagenes to file by factorizer and k
       for each factorizer 
          plot score vs k

@@

It is fundamental to our approach that metagenes determined on the basis of one cohort of data can be used to generate metasamples for a different cohort.  In the exposition below we refer to these as the \emph{training} and \emph{validation} cohorts respectively.

@@

% \subsection{Investigation of optimal parameters}

\subsection{Cluster coherence analysis results}

@@

Gene enrichment analysis against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term.   
To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF-go-lineage} to \ref{fig-PCA-1-3-go-lineage}.  
The contributing candidate genes for all significantly enriched terms are shown beneath each figure.  
(Ideally, the specific genes would be listed alongside each term, but this is difficult to achieve graphically).   
High-level terms at depth less than 3 were removed, since these are generic (e.g. "regulation of biological process") and so uninteresting.   
One component -- ICA-4-of-5 -- was problematic in that 49 GO terms (with depth $\geq 3$) were identified as significant.  
For this component it was necessary to limit the graphic to the 12 terms having the largest number of associated candidate genes.   
The full table of terms for this component is available in appendix \ref{tab-ica-4-of-5-enrichment-table}.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.5, draft=false]{\plotspath{TCGA_OV_VST/GeneEnrichment/component_jaccard_heatmap.pdf}}
\end{center}
\caption{Heatmap of Jaccard similarities between the candidate genes identified by the 11 components.}
\label{fig-component-jaccard-similarities}
\end{figure}

@@

\begin{sidewaysfigure}[ht]
\begin{center}
\includegraphics[width=\textwidth, draft=false]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_4_of_ICA_5.png}} \\
\begin{spacing}{0.6}
{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_4_of_ICA_5.txt}}
\end{spacing}
\vspace{0.5cm}
\caption{Lineage maps of enriched Gene Ongology (GO) terms for component ICA-4-of-5. For this component it was necessary to additionally filter the enriched terms -- see text.}
\label{fig-ICA-4-go-lineage}
\end{center}
\end{sidewaysfigure}

@@

\subsection{Metasample heatmap clustering results}

As with survival analysis, metasamples were derived from the metagenes factorized from the TCGA dataset, then applied to both TCGA and AOCS.  The resulting heatmaps are shown in figures \ref{fig-heatmap-TCGA-TCGA} and \ref{fig-heatmap-TCGA-OACS} respectively.   Metadata was available for the AOCS dataset, and this is presented in colour coded columns on the left of figure \ref{fig-heatmap-TCGA-OACS}.   The meanings of each column are as follows:
\begin{description}
\item[WGD]: Whole genome doubling -- a marker of genome instability.
\item[Cellularity]: proportion of cells belonging to the tumour (as opposed to surrounding normal tissue).
\item[HR Detect]: A predictor of homologous repair deficiency based on established mutational features \cite{Ewing2020}.
\item[Mutational Load]: A measure of the total number of mutations present in the tumour genome.
\item[CNV Load]: Copy-number variation, i.e. deviation from the normal diploid cell compliment.
% \item[Donor survival time]: (should probably remove).
\end{description}

>>>>

\documentclass[draft, tikz, 12pt,a4paper,oneside,fleqn]{article}

\usepackage[margin=1.5cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{setspace}
\usepackage{enumitem}
\setlist[description]{leftmargin=\parindent,labelindent=\parindent}

@@

\title{Disentangling patterns of gene expression in high grade serous ovarian cancer by matrix factorization}
\author{}

@@

%%%%%% Assesment Title Page
\begin{center}
\begin{center}
\includegraphics[scale=0.3]{images/UoE_SBO_logo.png}
\end{center}

@@

Disssertation Supervisor:  Dr. Ailith Ewing
\vspace{0.2in}
\end{center}
\end{mdframed}
\end{center}

@@

\begin{abstract}
Here is the abstract
\end{abstract}
\newpage

\section{Introduction}

\subsection{Epithelial Ovarian Cancer}

Ovarian cancer (OC) is a heterogeneous disease, presenting with a wide range of pathologies and molecular characteristics .  The World Health Organisation (WHO) sets out five subtypes of epithelial ovarian carcinomas (EOC) grouped into two main types\cite{Kossai2018, Lisio2019}:

\begin{description}
\item[Type I] are low grade with generally good prognosis, with two subtypes:
\begin{description}
\item[Low grade serous] affecting the epithelial membrane which secretes serous fluid), $< 5$\% of cases
\item[Mucosal] affecting membrane rich mucous glands, 2 - 3\% 
\item[Clear cell] relating to the presence of clear cells in histology, 5 - 10\%.
\end{description}
\item[Type II] are more aggressive (rapidly growing), typically carrying P53 mutations with defects in mechanisms of DNA repair. Three subtypes:
\begin{description}
\item[\emph{High grade serous OC (HGSOC)}] affects the epithelium of the ovaries and fallopian tubes.  It is the most common, accounting for 70\% of cases, and most serious of the OC subtypes and the focus of this dissertation.
\item[Endometrial] affecting the the inner lining of the uterus, 10 \% of cases.
\end{description}
\end{description}

HGSOC typically occur in older patients and are diagnosed at a late stage.   Histology of HGSOC is similar to the low grade subtype, but develops along distinct molecular pathways.

\emph{Treatment options for HGSOC...}

\subsection{Patterns of gene expression in HGSOC}

Identifying the patterns -- or ``signatures'' -- of gene expression has been an active research area, for the purposes of understanding the carcinogenesis pathway, subtype classification, predicting survival prognosis, predicting treatment response and understanding the causative mechanism of risk factors; these are clearly related, however.

\subsubsection{Carcinogenesis pathways including homologous repair deficiency
}

Studying the combined landscape of chromosomal rearrangements, driver mutations (particularly TP53, BRCA1/2), methylation and gene expression profiles has lead to a better understanding of the molecular events involved in the progression HGSOC and the development of chemoresistance \cite{Patch2015}.  The sequence of events is complex, however, and not easily summarised here.   A similar integrative / multiplatform approach is taken in \cite{He2017a} and \cite{Hoadley2014}.

\subsubsection{Subtype classification}

A relatively early (2003) expression microarray based work demonstrated clustering into for example lymphocyte related genes (IGH3, IGKC, IGLJ3), extracellular matrix/stromal related (COL11A1, Col3A1, MMP2, SPARC, RBBP1) and six others \cite{Schaner2003}.

Disruption of the PI3K-AKT signalling pathway is known to be significant in several cancers, due to its role in regulating apoptosis.  
Espinosa \etal\ \cite{Espinosa2011} used unsupervised clustering techniques to investigate expression of 22 genes of the pathway.  
They found that HGSOC cases formed two separate clusters; the cluster with high expression of CASP3, XIAP, NFKB1, FAS and GSK3B was linked with better outcomes.

\subsubsection{Treatment response and survival}

As we have seen in section \ref{sec-epithelial-ovarian-cancer}, EOC subtypes are closely linked with survival, however much research is directed at finding direct links between expression and survival.

The prognostic value of expression signatures is demonstrated by the ``Classification of Ovarian Cancer" (CLOVAR) system \cite{Verhaak2013}.  
Single-sample gene set enrichment analysis (ssGSEA) to obtain scores for previously reported expression signatures of differentiated, immunoreative, mesenchymal and proliferative.  
A survival outcome prediction model based on these signatures gave good survival stratification (likelihood ratio 10.8 on a validation dataset) which improved to 15.7 when augmented with BRCA1/2 mutation status and other clinical post-operative details.

An explicit supervised learning approach can be taken, as in Berchuck \etal, where a hybrid decision tree plus linear discriminant model demonstrated survival stratification in both cross-validation and external validation experiments, based on microarray expression data.   
They found expression of CSFT3, ABCD3, MAL and APMCF1 to be the most influential; in fact the decision tree was almost entirely driven by the first two of these

Finding a clear correlation between gene expression patterns and survival is not straight forward.  
One substantial meta-analysis applied 16 previously published gene sets, to two studies from GEO (199 samples) which included survival information, and found no significant predictive power \cite{Fekete2012a}.  
Although, within the same study, the authors did identify predictive genes, in particular SNGC, MAPT, ESR2 and PGR.

Expression of a single gene -- CD38, which codes for a transmembrane glycoprotein -- has recently been shown to have survival prognostic value on its own \cite{Zhu2020}.
CD38 has a role in the breakdown of nicotinamide adenine dinucleotide ($\mbox{NAD}^+$), and is also a cell marker of lymphocytes.
CD38 is over expressed in myloma cells, thence its importance as a prognostic marker and a potential drug treatment target.  
It is worth mentioning that the analysis in this paper made use of the Gene Expression Profiling Interactive Analysis (GEPIA) database, which conveniently supports survival analysis with respect of any given gene, against a range of TCGA gene expression databases.

\subsubsection{Association with other risk factors}

Obesity has been shown to negatively impact prognosis in ovarian cancer, possibly through difficulties in matching chemotherapy dose with patient BMI.  \cite{Au-Yeung2014}  A more direct molecular link is proposed by Cuello \etal\ \cite{Cuello2018}.  They used NMF based clustering (see section \ref{sec-nmf}) to demonstrate that expression of genes linked with obesity and lipid metabolism impart poorer progression free survival, independent of known HGSOC driver gene expression.  The link is complex however, there been many confounding non-molecular based reasons for a link between high BMI and HGSOC survival, including later diagnosis and compromised operability.  Some studies (e.g. \cite{Hew2013}) conclude no overall link at all.

Note that several of the studies reviewed here use data from the Australian Ovarian Cancer Study (AOCS): \cite{Patch2015,Ewing2020,Cuello2018,Au-Yeung2014}.

%Stratification of HGSOC by immunohistochemistry is addresses Kamble \etal. \cite{Kamble2019}.  They identify a panel of six biomarkers -- TCF21, E-cadherin, PARP1, Slug and AnnexinA2.   These markers allow stratification into categories including Mesenchymal-to-epithelial transition, Homologous recombination repair and Epithelial-to-mesenchymal transition.  This, they argue, allows selection of class-specific inhibitor drugs such as Olaparib and Rucaparib.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{Sompairac_ICA_NMF_PCA.png}
\end{center}
\caption{Illustrating the differing results of component extraction by PCA, ICA and NMF.  From \cite{Sompairac2019}}
\label{fig-pca-ica-nmp-explanation}
\end{figure}

@@

In this work I wish to treat and discuss the three factorization methods in a unified way.  Thus, for the remainder of this dissertation the following notational conventions are adopted:
\begin{description}
\item[Expression matrix:] $X$ $(M \times N)$
\item[Metagene matrix:] $W$ $(M \times K)$
\item[Metasample matrix:] $H$ $(K \times N)$
\end{description}
where $N$ is the number of patients (or samples), $M$ is the number of genes and $K$ is the factorization rank, i.e. the number of components or factors.  Thus the factorization is written
\begin{equation}
X \approx W H
\end{equation}
Adding subscripts to indicate the row,column orientation of the matrices:
\begin{equation}
X_{M,N} \approx W_{M,K} H_{K,N}
\end{equation}
as illustrated in figure \ref{fig-matrix-factorization} (a).

@@

\subsection{Research goals}
This work is concerned with analysing the patters gene expression in HGSOC by dimensionality reduction techniques of matrix factorization, and relating these patterns to clinical and genomic features.

The following research questions were set out early in the project. 
\begin{enumerate}
\item Which are the influential genes in HGSOC according to gene expression data?   Do these confirm published results?
\item Which of PCA, ICA and NMF is best suited to this analysis?
\item For identified genes, what is the underlying biology?
\item Can we discriminate subtypes of HGSOC from expression data?
\item Is there value in incorporating somatic and germline genomic data?
\end{enumerate}

@@

Our overall approach is to use \emph{unsupervised} machine learning methods to represent gene expression data in a small number of features, then to study whether these features correlate with clinical and biological variables.  This is appropriate in our case, since of the two datasets available (described below), one (TCGA) has relatively large $n$ but little available metadata, while the other (AOCS) is much smaller but has more useful metadata of genomic features.   This motivates an approach of unsupervised learning on the TCGA dataset.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.7]{methodlology_overview.pdf}
\end{center}
\caption{Overview of methodology as information flow.  The diagramatic convention here is that at each stage, both datasets -- TCGA and AOCS -- are processed separately by the given algorithm.  The exception is Gene Set Intersection, which involves interaction between the datasets.}
\label{fig-method-overview}
\end{figure}
In outline our method is follows:
\begin{enumerate}
\item
Identification of a consistent gene set between the TCGA and AOCS datasets.
\item
Unsupervised metagene extraction by matrix factorization methods : NMF, ICA and PCA.
\item
Metagene selection for robustness by k-means clustering of bootstrap resampled factorizations evaluated by silhouette score, based on the TCGA dataset.
\item
Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO).
\item
Computation by least squares optimisation of metasamples associated with with each TCGA derived metagene against TCGA and AOCS datasets.
\item
Alignment of available metadata with computed metasamples, for TCGA and AOCS.
\item 
Survival analysis on TCGA and AOCS to investigate relationship between metasamples and patient survival.
\item
Clustered heatmaps to visualise relationships between metasamples, patients and metadata.
\end{enumerate}

@@

       Ws_reduced = PCA(Ws, rank=20)
       plot t-SNE(Ws_reduced)
       clustering = k-means clustering(Ws_reduced, n_clusters=k)
       score[factorizer, k] = silhouette score(clustering)	  
       median_metagenes = calculate medians for k clusters of metagenes
       save median_metagenes to file by factorizer and k
       for each factorizer:
          plot score vs k

@@

It is fundamental to our approach that \emph{metagenes} determined on the basis of one cohort of data can be used to generate \emph{metasamples} for a different cohort.  In the exposition below we refer to these as the \emph{training} and \emph{validation} cohorts respectively.

@@

% \subsection{Investigation of optimal parameters}
\subsection{Cluster coherence analysis results}

@@

Gene enrichment analysis against the GO results in, for each of the 11 metagenes, a table of GO terms with a list of the candidate genes which have an annotated association to that term.   
To take full advantage of the hierarchical nature of the GO, results are presented as lineage maps in figures \ref{fig-NMF-go-lineage} to \ref{fig-PCA-1-3-genesiso-lineage}.  
The contributing candidate genes for all significantly enriched terms are shown beneath each figure.  
(Ideally, the specific genes would be listed alongside each term, but this is difficult to achieve graphically).   
High-level terms at depth less than 3 were removed, since these are generic (e.g. "regulation of biological process") and so uninteresting.   
One component -- ICA-4-of-5 -- was problematic in that 49 GO terms (with depth $\geq 3$) were identified as significant.  
For this component it was necessary to limit the graphic to the 12 terms having the largest number of associated candidate genes.   
The full table of terms for this component is available in appendix \ref{tab-ica-4-of-5-enrichment-table}.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{\plotspath{TCGA_OV_VST/GeneEnrichment/component_jaccard_heatmap.pdf}}
\end{center}
\caption{Heatmap of Jaccard similarities between the candidate genes identified by the 11 components.}
\label{fig-component-jaccard-similarities}
\end{figure}

@@

\begin{sidewaysfigure}[ht]
\begin{center}
\includegraphics[width=\textwidth]{\plotspath{TCGA_OV_VST/GeneEnrichment/go_lineage_4_of_ICA_5.png}} \\
\begin{spacing}{0.6}
{\small Genes: \tiny \input{/home/ipoole/Documents/gitrepos/HgsocTromics/Plots/TCGA_OV_VST/GeneEnrichment/go_gene_set_4_of_ICA_5.txt}}
\end{spacing}
\vspace{0.5cm}
\caption{Lineage maps of enriched Gene Ongology (GO) terms for component ICA-4-of-5. For this component it was necessary to additionally filter the enriched terms -- see text.}
\label{fig-ICA-4-go-lineage}
\end{center}
\end{sidewaysfigure}

@@

\subsection{Metasample heatmap clustering results}
As with survival analysis, metasamples were derived from the metagenes factorized from the TCGA dataset, then applied to both TCGA and AOCS.  The resulting heatmaps are shown in figures \ref{fig-heatmap-TCGA-TCGA} and \ref{fig-heatmap-TCGA-OACS} respectively.   Metadata was available for the AOCS dataset, and this is presented in colour coded columns on the left of figure \ref{fig-heatmap-TCGA-OACS}.   The meanings of each column are as follows:
\begin{description}
\item[WGD]: Whole genome doubling -- a marker of genome instability.
\item[Cellularity]: proportion of cells belonging to the tumour (as opposed to surrounding normal tissue).
\item[HR Detect]: A predictor of homologous repair deficiency based on established mutational features \cite{Ewing2020}.
\item[Mutational Load]: A measure of the total number of mutations present in the tumour genome.
\item[CNV Load]: Copy-number variation, i.e. deviation from the normal diploid cell compliment.
% \item[Donor survival time]: (should probably remove).
\end{description}
